Estimating duration using survival analysis

### 2024-03-25 cis-perfect-testing.tex
proof

The CIS samples individuals less frequently, but has a large sample size (\numprint{4800} detected infection episodes) and up to years of follow-up.
Therefore, it is suitable for estimating the tail of the duration distribution.
In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming the test ??? has perfect sensitivity and specificity, {\it i.e.} there are no false negatives or positives.
Formally, I aim to estimate the survival function $S(t)$ defined in \cref{inc-prev:sec:definitions}.

In \cref{perf-test:sec:problem}, I describe the challenges analysing the data presents.
This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
In \cref{perf-test:sec:parameters-priors}, I consider how to parameterize the survival function and choose its priors, including vague and smoothing priors as well as a prior constructed to include the information from \cref{E-ATACCC}.
I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they perform well.
Finally, I discuss the results in \cref{perf-test:sec:discussion}.

@@

I use Bayesian inference for two reasons.
First, the frequentist properties (\eg consistency and convergence rate) of proposed estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
Second, a prior can include information from previous studies, such as \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard, which is biologically sensible, and can overcome issues caused by the undetected episodes ~\autocite{caoBias}.

@@

In this section, I develop a computationally feasible statistical model accounting for double interval censoring and undetected episodes.
In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
Then, in \cref{perf-test:sec:episode-types}, I apply the framework of \textcite{heiseyModelling} which classifies possible episode beginning and end time into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred and been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
The number of such episodes in each class has a multinomial distribution conditional on the latent total number of infections (including undetected infections).

@@

I consider the cohort of $\Ncis$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
To be considered, an individual must have had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
I index these individuals by $i = 1, \dots, \Ncis$.

@@

For a detected episode $j$, five pieces of data are observed (see \cref{E-biology-data:sec:cis-episodes} for how these are constructed).
The individual in which the episode occurs, $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
Over the short calendar period considered, each individual has zero or one detected episodes.
Let $i \in \set{D}$ indicate individual $i$ has a detected episode.

In addition, the individuals in the cohort experienced an unknown number of undetected infections, $\nnodet$.
The total number of infections in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.

Any infection episode, whether detected or not, is fully characterised for my purposes by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ denotes the individual the episode occurred in.
The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
For all $j$, the exact values $b_j$ and $e_j$ are latent.
If $j$ is undetected then $i(j)$ is also latent and nothing is known about $b_j$ or $e_j$.
If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
These latter two properties are what make the data doubly interval censored.

@@

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$), shown in \cref{perf-test:fig:partitionSpace}.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, specified in \cref{perf-test:sec:parameters-priors}.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$; in this case individual $i$ had $n_{ia} = 0$ admissible episodes.
  Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
  The admissible region is the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
  This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
  The shape of this region accounts for the doubly interval censored nature of the data.
  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these occur in individual $i$.
  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, indicating combinations of episode beginning and end times which do not span a testing time plus episodes beginning before the individual's enrolment.
  The shape of this region accounts for the pattern of possible undetected infections.
  Conditional on the episode occurring in individual $i$, the episode is undetected with probability $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
\item
  Inadmissible episodes, $\beta_i$, which would be detected (if they occur) but were not.
  Therefore, we know $n_{i\bar{a}} = 0$ of these occurred in individual $i$.
  These are all remaining episodes not in the previous sets.
  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$.
\end{itemize}

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption[Admissible, inadmissible, and undetected regions]{%
  The regions for individual $i$.
  $i$'s first test was at time 0 and negative, with subsequent negative tests at times 7, 14, and 56, and positive tests at times 21 and 28.
  Inadmissible regions are unshaded.
  The black (impossible) region covers the cases in which recovery occurs prior to infection.
  \label{perf-test:fig:partitionSpace}
}
\end{figure}

These three classes span all possible episodes and are mutually exclusive.
Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0$.
Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.

% Any episode $j$ independently belongs to one of these classes.
The probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$), are conditional on the episode occurring in individual $i$.
I assume the individual in which each episode occurs is iid uniform across the cohort.
Therefore, these probabilities are multiplied by $1/\Ncis$ to form the unconditional probabilities.
For example, the probability of an episode being in class $\alpha_i$ is $p_{ia} / \Ncis$.

Each episode occurs in one of the classes independently and with the same probability.
Therefore:
\begin{align}
  \begin{bmatrix}
    n_{1a} \\ n_{1u} \\ n_{1\bar{a}} \\ \vdots \\ n_{\Ncis,a} \\ n_{\Ncis,u} \\ n_{\Ncis,\bar{a}}
  \end{bmatrix}
  \mid \ntot
  =
  \MN\left(
    \ntot,
    \frac{1}{\Ncis}
    \begin{bmatrix}
      p_{1a} \\ p_{1u} \\ p_{1\bar{a}} \\ \vdots \\ p_{\Ncis,a} \\ p_{\Ncis,u} \\ p_{\Ncis,\bar{a}}
    \end{bmatrix}
  \right)
  \label{perf-test:eq:multinomial}
\end{align}

In forming the region $\alpha_i$, I use the test outcomes.
This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
However, an alternative considering each combination of infection and recovery times as its own cell is possible.
The new cells would align with the inadmissible region.
Since each of these cells have an observed count of zero, they can be combined to give the same likelihood.
That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
Therefore, I present this version for clarity.

The undetected classes cannot be distinguished.
Therefore, I combine these classes into a single class.
The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability parameter $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.

@@

Next, I derive $p_{ia}$.
From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.
For such an individual, let $j(i)$ denote their detected episode.

@@

When constructing the prior, two aspects need consideration.
Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
That is, the prior used in this analysis should have correlation between $\lambda_t$ and $\lambda_{t'}$.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior to include this prior information into the CIS analysis two steps.

@@

\ifSubfilesClassLoaded{
  \listoftodos
}{}
\end{document}

>>>>

The CIS samples individuals less frequently, but has a large sample size (\numprint{4800} detected infection episodes) and up to years of follow-up.
Therefore, it is suitable for estimating the tail of the duration distribution.
In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming the test has perfect sensitivity and specificity, {\it i.e.} there are no false negatives or positives.
Formally, I aim to estimate the survival function $S(t)$ defined in \cref{inc-prev:sec:definitions}.

In \cref{perf-test:sec:problem}, I describe the challenges analysing the data presents.
This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
In \cref{perf-test:sec:parameters-priors}, I consider how to parameterize the survival function and choose its priors, including vague and smoothing priors as well as a prior constructed to include the information from \cref{E-ATACCC}.
I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they perform well.
Finally, I discuss the results in \c{perf-test:sec:discussion}.

@@

I use Bayesian inference for two reasons.
First, the frequentist properties (\eg consistency and convergence rate) of existing estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
Second, a prior can include information from previous studies, such as \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard, which is biologically sensible, and can overcome issues caused by the undetected episodes ~\autocite{caoBias}.

@@

In this section, I develop a computationally feasible Bayesian statistical model accounting for double interval censoring and undetected episodes.
In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
Then, in \cref{perf-test:sec:episode-types}, I adapt the framework of \textcite{heiseyModelling} which classifies possible episode beginning and end time into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred and been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
The number of such episodes in each class has a multinomial distribution conditional on the latent total number of infections (including undetected infections).

@@

I consider the cohort of $\Ncis = \numprint{437590}$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
To be considered, an individual must have had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
I index these individuals by $i = 1, \dots, \Ncis$.

@@

For a detected episode $j$, five pieces of data are observed (see \cref{E-biology-data:sec:cis-episodes} for how these are constructed).
The individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
Over the short calendar period considered, each individual has zero or one detected episodes.
Let $i \in \set{D}$ indicate individual $i$ has a detected episode and denote $j(i)$ as the detected episode in individual $i$; $j(i)$ is only defined for $i \in \set{D}$.

In addition, the individuals in the cohort experienced a total of $\nnodet$ undetected infection episodes.
$\nnodet$ is a latent quantity.
The total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.

Any infection episode, whether detected or not, is fully characterised for my purposes by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ is the individual the episode occurred in.
The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
For all $j$, the exact values $b_j$ and $e_j$ are latent.
If $j$ is undetected then $i(j)$ is also latent and nothing is known about $b_j$ or $e_j$.
If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
These latter two properties are what make the data doubly interval censored.

@@

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$), shown in \cref{perf-test:fig:partitionSpace}.
Each episode, whether detected or not, must fall into one of these three classes with a probability that depends on the parameters of the survival function, $\vec{\theta}$, specified in \cref{perf-test:sec:parameters-priors}.
The regions are defined for each individual $i$.

\begin{itemize}
\item
  The region of admissible episodes for individual $i$, $\alpha_i$.
  An admissible episode has an infection and recovery time consistent with the data observed.
  Let $n_{ia}$ be the number of admissible episodes for $i$.
  $\alpha_i$ is empty if $i$ had no detected episodes (\ie $i \notin \set{D}$); in this case the observation is $n_{ia} = 0$.
  Otherwise, if $i \in \set{D}$, then $n_{ia} =1$.
  The admissible region is the intersection of the intervals $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$ and $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
  This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
  The shape of this region accounts for the doubly interval censored nature of the data.
  Let $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \alpha_i$ given that $j$ occurs in individual $i$.
  \label{perf-test:def:admissible}
\item
  The region of undetected episode for individual $i$, $\Omega_i$.
  Let $n_{iu}$ be the (latent) number of undetected episodes for $i$.
  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, indicating combinations of episode beginning and end times which do not span a testing time plus episodes beginning before the individual's enrolment.
  Let $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \Omega_i$ given that $j$ occurs in individual $i$.
\item
  The region of inadmissible episodes, $\beta_i$ for individual $i$.
  An inadmissible episode is one which would have been detected had it occurred, but the data rule out having occurred.
  Let $n_{i\bar{a}}$ be the number of inadmissible episodes for $i$, which is observed as $n_{i\bar{a}} = 0$.
  These are all remaining (a priori possible) episodes not in the previous sets.
  Let $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \beta_i$ given that $j$ occurs in individual $i$.
\end{itemize}

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption[Admissible, inadmissible, and undetected regions]{%
  The regions for individual $i$.
  $i$'s first test was at time 0 and negative, with subsequent negative tests at times 7, 14, and 56, and positive tests at times 21 and 28.
  Inadmissible regions are unshaded.
  The grey (impossible) region covers the cases in which recovery occurs prior to infection.
  \label{perf-test:fig:partitionSpace}
}
\end{figure}

These three regions span all possible episodes and are mutually exclusive.
Hence the following relations hold:
\begin{align}
\ntot
&= \ndet + \nnodet + \sum_{i=1}^{\Ncis} n_{i\bar{a}} \\
&= \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0
\intertext{and, for all $i$,}
p_{ia} + p_{iu} + p_{i\bar{a}} &= 1.
\end{align}

I will use these regions to form the data likelihood, conditional on $\ntot$.
The definition of the admissible region relies on the test results.
However the likelihood should be conditional only on $\vec{\theta}$, $\ntot$, and the test schedules.
Therefore, this is not strictly correct.
This issue can be rectified be considering the inadmissible region as many subregions each with an observed count of zero (sketched in \cref{perf-test:sec:alt-likelihood}).
Count zero regions do not contribute to the probability meaning the result is the same.
A formalisation of this approach requires much more complex notation for the same result; therefore, I prefer this presentation.

% Any episode $j$ independently belongs to one of these classes.
The probabilities $p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$ are conditional on the episode occurring in individual $i$.
I assume the individual in which an episode occurs is iid uniform across the cohort, that is $\prob(i(j) = i) = 1/\Ncis$.
Therefore, the conditional probabilities are multiplied by $1/\Ncis$ to form the unconditional probabilities.
For example, the probability of an episode being in class $\alpha_i$ is $p_{ia} / \Ncis$.

An episode occurs in one of the regions independently and with the same probability.
Therefore:
\begin{align}
  \begin{bmatrix}
    n_{1a} \\ n_{1u} \\ n_{1\bar{a}} \\ \vdots \\ n_{\Ncis,a} \\ n_{\Ncis,u} \\ n_{\Ncis,\bar{a}}
  \end{bmatrix}
  \mid \ntot, \vec{\theta}
  \dist
  \MN\left(
    \ntot,
    \frac{1}{\Ncis}
    \begin{bmatrix}
      p_{1a} \\ p_{1u} \\ p_{1\bar{a}} \\ \vdots \\ p_{\Ncis,a} \\ p_{\Ncis,u} \\ p_{\Ncis,\bar{a}}
    \end{bmatrix}
  \right)
  \label{perf-test:eq:multinomial}
\end{align}

The undetected regions cannot be distinguished.
Therefore, I combine these regions into a single cell within the multinomial distribution.
The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability parameter $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.

@@

Next, I derive $p_{ia}$.
From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.

@@

When constructing the prior, two aspects need consideration.
Firstly, the model structure from \cref{E-ATACCC} leads to the posterior distribution on the hazards having a positive correlation between $\lambda_t$ and $\lambda_{t'}$, especially when $t$ and $t'$ have a small difference.
The prior used in this analysis should preserve this correlation.
Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
I form the prior to include this prior information into the CIS analysis two steps.

@@

\ifSubfilesClassLoaded{
  \appendix
  \subfile{cis-perf-new-ll}
  \listoftodos
}{}
\end{document}
